---
title: "ACE Conference Presentation"
output:
  pdf_document: default
  html_document: default
---
Here are the needed libraries for this tutorial

```{r setup, include=FALSE}
#install.packages('tidyverse')
#install.packages('devtools')
#install.packages("usmap")
#install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
#install.packages("sf")
#install.packages("raster")
#install.packages("dplyr")
#install.packages("spData")
#install.packages("spdep")
#Note that you will have to install spDataLarge after installing spData
#install.packages('spDataLarge', repos='https://nowosad.github.io/drat/', type='source')
#install.packages("rgdal")
#install.packages("sp")
#install.packages("tableone")
#install.packages("MASS")
#install.packages("BayesLogit")
#install.packages("FRK")
#install.packages("fields")
#install.packages("mpath")
#install.packages("SemiPar")
#install.packages("data.table")
#install.packages("stringr")
#install.packages("car")
#install.packages("readxl")


library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(spdep)
library(tidyverse)
library(usmap)
library(ggplot2)
library(RColorBrewer)
library(INLA)
library(rgdal)
library(sp)
library(tableone)
library("MASS")
library("BayesLogit")
library("FRK")
library("fields")
library("mpath")
library("SemiPar")
library("data.table")
library("stringr")
library("car")
library("readxl")

```

## Research Question and Data file

Research Question: What are the sociodemographic factors associated with West Nile virus rates in Pennsylvania?

Let's load the data:

```{r cars}
#change setwd to the folder the downloaded files are located
setwd("C:/Users/John/Desktop/aphaworkshop/aceconference")

#load the West Nile Virus data
case.data = read.csv("PA.WN.covariates.csv", header = T)

head(case.data)


```

#Calculate Expected Counts: Offset Variable
(Total cases/Total Population) * County Population
```{r}
#generating the expected counts
case.data$EC = sum(case.data$TotalCases)/sum(case.data$tot_pop)*case.data$tot_pop
```

## Desriptive Statistics

First Task: Let's visualize the distribution of cases of West Nile virus: "TotalCases"

```{r pressure, echo=FALSE}

hist(case.data$TotalCases)

table(case.data$TotalCases, useNA = "always")

```

Let's check these distribution on a choropleth

First, let's check a simple choropleth showing the states and counties in the US

```{r}
plot_usmap(regions = "states") + 
  labs(title = "U.S. States",
       subtitle = "Map of US States.") + 
  theme(panel.background=element_blank())


plot_usmap(regions = "counties") + 
  labs(title = "U.S. counties",
       subtitle = "Map of US Counties") + 
  theme(panel.background=element_blank())
```


Our data is focused on Pennsylvania. Let's select PA data (using "include" syntax) at the county level

```{r}
plot_usmap(regions = "counties", include = c("PA")) +
  labs(title = "Pennsylvania") +
  theme(panel.background = element_rect(color = "black"))
```

Let's add the West Nile virus counts to the map

```{r}
#Quick Check: the identifier is "fips" NOT "FIPS" - case sensitive

#Let's plot

plot_usmap(data = case.data, values = "TotalCases", include = "PA", color = "blue") + 
  scale_fill_continuous(low = "white", high = "blue", name = "Cases", label = scales::comma) + 
  labs(title = "West Nile Virus Cases in Pennysylvania", subtitle = "Cases per County for PA counties") +
  theme(legend.position = "right")


```
We can generate categories to better visualize the distribution

We can create "casecat" and have 4 categories
0 count, 1 count, 2-5 counts, 6 or more counts


```{r}
table(case.data$TotalCases)


case.data$casecat = ifelse(case.data$TotalCases==0,0,
                           ifelse(case.data$TotalCases==1, 1,
                                  ifelse(case.data$TotalCases>=2 & case.data$TotalCases<=5,2,
                                         ifelse(case.data$TotalCases>5, 3, NA))))


case.data$casecat <- ordered(case.data$casecat,
                             levels = c(0,1,2,3),
                             labels = c("0", "1", "2-5",">5"))

table(case.data$casecat)
```

Let's map these fatal crash categories
we change continous syntax to discrete syntax in ggplot2
A good quick resource: www.ggplot2tutor.com/scales/scale_fill_discrete

```{r}
#without a theme 
plot_usmap(include = "PA", data = case.data, values = "casecat") + 
  scale_fill_discrete(name = "West Nile Virus Counts", h= c(0,360) + 15, c=100, l=65, h.start =0, direction=1, na.value="grey50", aesthetics ="fill") +
  labs(title = "West Nile Virsu Counts in Pennsylvania", subtitle = "West Nile Virus in PA")

#with a theme

plot_usmap(include = "PA", data = case.data, values = "casecat") + 
  scale_fill_discrete(name = "West Nile Virus Counts", h= c(0,360) + 15, c=100, l=65, h.start =0, direction=1, na.value="grey50", aesthetics ="fill") +
  labs(title = "West Nile Virus Counts in Pennsylvania", subtitle = "West Nile Virus in PA") + 
  theme(panel.background = element_rect(colour = "black", fill = "light blue")) + theme(legend.position = "right")



```


# Fatality Rates

Case counts are not very informative. Let's have some standardized measures like Crude Rates per 100,000 population

```{r}

case.data$caserate = round(case.data$TotalCases/case.data$tot_pop*100000, 0)

hist(case.data$caserate)

table(case.data$caserate, useNA = "always")

```


```{r}


ruralcases <- case.data$TotalCases[case.data$ruca2gp == "Low comuting micropolitan, Rural and Small towns"]
urbancases <- case.data$TotalCases[case.data$ruca2gp == "Metropolitan/Urban"]


num_breaks_rural <- 5  # Number of bins for rural
num_breaks_urban <- 10  # Number of bins for urban


hist(ruralcases, breaks = num_breaks_rural, col = "blue", main = "Histogram for West Nile Virus in Rural Areas", xlab = "Counts")


hist(urbancases, breaks = num_breaks_urban, col = "red", main = "Histogram for West Nile Virus in Urban Areas", xlab = "Counts")




#what if we want to keep the x-axis of the same length
# data_range <- range(c(ruralcases, urbancases))
# num_breaks <- 5
# 
# #Now the histogram
# 
# hist(ruralcases, breaks = seq(data_range[1], data_range[2], length.out = num_breaks + 1),  xlim = data_range, col = "blue", main = "Histogram for West Nile Virus in Rural Areas", xlab = "Counts")
# 
# 
# hist(urbancases, breaks = seq(data_range[1], data_range[2], length.out = num_breaks + 1),  xlim = data_range, col = "red", main = "Histogram for West Nile Virus in Urban Areas", xlab = "Counts")

```



Let's generate categories and visualize the categories

```{r}
table(case.data$caserate)


case.data$ratecat = ifelse(case.data$caserate==0,0,
                          ifelse(case.data$caserate==1,1,
                           ifelse(case.data$caserate==2,2,
                                  ifelse(case.data$caserate>=2,3,NA))))


case.data$ratecat <- ordered(case.data$ratecat,
                             levels = c(0,1,2,3),
                             labels = c("0", "1", "2", "3 or more"))

table(case.data$ratecat)

#replace scale_fill_discrete with scale_colour_grey
plot_usmap(include = "PA", data = case.data, values = "ratecat") + scale_colour_grey(start=0.7, end= 0) + labs(title = "West Nile Virus Counts in Pennsylvania", subtitle = "West Nile Virus in PA") + theme(panel.background = element_rect(colour = "black", fill = "light blue")) + theme(legend.position = "right")



```

#Visualizing the Standardized Mortality Rate
```{r}
#calculating the Standardized rate ratio (from SMR): Observed/expected
case.data$smr = round(case.data$TotalCases/case.data$EC,2)

hist(case.data$smr)

plot_usmap(data = case.data, values = "smr", include = "PA", color = "blue") + 
  scale_fill_continuous(low = "white", high = "blue", name = "SMR", label = scales::comma) + 
  labs(title = "Standardized Rate Ratio", subtitle = "RR for Pennysylvania Counties") +
  theme(legend.position = "right")

table(case.data$smr)

case.data$smrcat = ifelse(case.data$smr<=0.50,0,
                         ifelse(case.data$smr>=0.51 & case.data$smr<1.00,1,
                                ifelse(case.data$smr>=1.00 & case.data$smr<2.00,2,
                                       ifelse(case.data$smr>=2.00,3, NA))))


case.data$smrcat <- ordered(case.data$smrcat,
                           levels = c(0,1,2,3),
                           labels = c("0-0.50", "0.51-0.99", "1.00-1.99", ">=2.00"))

plot_usmap(include = "PA", data = case.data, values = "smrcat") + 
  scale_fill_discrete(name = "SMR", h= c(0,360) + 15, c=100, l=65, h.start =0, direction=1, na.value="grey50", aesthetics ="fill") +
  labs(title = "West Nile Virus Ratio in Pennyslyvania", subtitle = "Standardized Rate Ratio") + 
  theme(panel.background = element_rect(colour = "black", fill = "light blue")) + theme(legend.position = "right")


```
#Predictor Variables
Let's address the research question by listing some county demographic and health characteristics as predictors
1. Age 65 year and older: years65
2. Proportion of Males: Male
3. African American Proportion: Blackpct
4. Poverty Rates: Poverty
5. Rurality/Urbanicity: Binary variable - Rural/Urban
6. Computing expected death counts as an offset variable

```{r}


tab1 <- CreateTableOne (vars = c ("years65", "Male", "Blackpct",
                                  "Poverty", "EC"),
                        data = case.data)
print(tab1, showAllLevels = TRUE)
summary(tab1)

table(case.data$ruca2gp)

```
#Regression Analysis with Negative Binomial 

We run the regression analysis with INLA
First, we will specify the formula
Then insert it into the INLA syntax
More information and tutorials on INLA: https://www.r-inla.org/examples-tutorials

INLA: 
computationally efficient - fast alternative to MCMC
uses the Stochastic Partial Differentiation Equation (SPDE) to estimate spatial autocorrelation of the data

```{r}
#setting priors
#https://becarioprecario.bitbucket.io/inla-gitbook/ch-priors.html

prior.fixed1 <- list(mean.intercept = 0, prec.intercept = 0.0001,
                    mean = c(0, 0, 0), prec = c(0.0001, 0.0001, 0.0001))
prec.prior1 <- list(prec = list(param = c(0.001, 0.001)))


formula1 <- TotalCases ~ years65 + Male + Blackpct + Poverty + ruca2gp

dat.inla <- inla(formula1, family='nbinomial', E=EC,
                 data=case.data,
                 control.family=list(link='log'),
                 control.fixed = prior.fixed1,
                 control.predictor=list(link=1, compute=TRUE),
                 control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE))

summary(dat.inla)
```



What is the Summary Here?

Can we generate the Rate Ratios?
```{r}
#Older Adults
YearsRatio <- exp(-0.135)
LCI <- exp(-0.254)
UCI <- exp(-0.020)

YearsRatio
LCI
UCI

```
Every unit increase in Older adult population in a county is associated with a 23% reduction in the rate of West Nile virus in a county (95% Credible Interval: 0.78 - 0.98)




#Spatial Regression
1. Read the shape file
```{r}
pa.tr <- readOGR("C:/Users/John/Desktop/aphaworkshop/aceconference/PaCounty2023_08.shp")
```

2. Create adjacency matrix
```{r}
pa.adj <- poly2nb(pa.tr)
```


3. Non-standardized and Row standardized matrices: Generating Weights
```{r}
W.pa <- nb2mat(pa.adj, style = "B") 
W.pa.rs <- nb2mat(pa.adj, style = "W") 

```


```{r}
#Will the model improved by adding spatial estimators?
#AIC
#BIC
#DIc
#Moran's I


nonspatialmodel = glm.nb(TotalCases ~ years65 + Male + Blackpct + Poverty + ruca2gp, data = case.data)

nonspatialmodel

wt2.listw <- nb2listw(pa.adj, style = "W", zero.policy = TRUE)

class(wt2.listw)

moran.test(residuals.glm(nonspatialmodel),wt2.listw)

```



4. Merge the variables in the case.data with the spatial data

```{r}

#First, let's create a 5-digit fips variable

pa.tr@data$stcode = "42"

pa.tr@data$fips = paste0(pa.tr@data$stcode,pa.tr@data$FIPS_COUNT)

#Next: let's merge by fips

pa.tr@data= merge(pa.tr@data, case.data, by = 'fips')

```



5. Create the model

```{r}

pa.tr@data$ID <- 1:nrow((pa.tr@data))

pa.formula <- TotalCases ~ years65 + Male + Blackpct + Poverty + ruca2gp + f(ID, model = "bym", graph = W.pa.rs)



#BYM = besag+iid; try out besag, besagproper, bym2 and check the WAIC and DIC
#more information on the sensitivity of the models: https://becarioprecario.bitbucket.io/inla-gitbook/ch-spatial.html 

```

5. INLA model

```{r}

pa.bym <- inla(pa.formula,
                      family='nbinomial',E=EC,  
               data = as.data.frame(pa.tr@data),
               control.family=list(link='log'),
               control.fixed = prior.fixed1,
               control.compute = list (dic = TRUE, waic = TRUE, cpo = TRUE, return.marginals.predictor=TRUE),
               control.predictor = list(link=1, compute = TRUE)
)


summary(pa.bym)


```
Let's interpret the results
```{r}
#Older Adults
Sp_OlderAdult_RateRatio <- exp(-0.134)
Sp_LCI <- exp(-0.255)
Sp_UCI <- exp(-0.017)

Sp_OlderAdult_RateRatio
Sp_LCI
Sp_UCI

```
A unit increase in the proportion of older adults in PA counties is associated with a 23% reduction in the West Nile virus rates. 
(95% Credible Interval: 0.77 - 0.98)

Non-Spatial Model:    WAIC:222.50    DIC:219.20
Spatial Model:        WAIC:220.64    DIC:218.45


#6. Adjusted fatal counts

```{r}
#https://www.paulamoraga.com/book-geospatial/sec-inla.html
#summary.fitted.values: obtained by transforming the linear predictors by the inverse of the link function

head(pa.bym$summary.fitted.values)

pa.tr@data$AdjCounts <- pa.bym$summary.fitted.values[, "mean"]
pa.tr$LL <- pa.bym$summary.fitted.values[, "0.025quant"]
pa.tr$UL <- pa.bym$summary.fitted.values[, "0.975quant"]


```


#8. Mapping Exceedance Probability and Adjusted Rates

 We have adjusted estimates. Lets visualize the results.
First, let's convert back to simple features (sf)
Next, we plot.


```{r}
pa.tr_sf <- st_as_sf(pa.tr)

hist(pa.tr_sf$AdjCounts)
table(pa.tr_sf$AdjCounts)
mean(pa.tr_sf$AdjCounts)

ggplot(pa.tr_sf) + geom_sf(aes(fill = AdjCounts)) +
  scale_fill_gradient2(
    midpoint = 0.7, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()


```

```{r}
pa.tr$exc1 <- sapply(pa.bym$marginals.fitted.values, FUN = function(marg){1 - inla.pmarginal(q = 1.0, marginal = marg)})

```




```{r}

#mapping Exceedance probability

#https://www.paulamoraga.com/book-geospatial/sec-inla.html
#marginals.fitted.values: posterior marginals of the fitted values obtained by transforming the linear predictors by the inverse of the link function.


head(pa.bym$summary.fitted.values)

pa.tr$AdjCounts <- pa.bym$summary.fitted.values[, "mean"]
pa.tr$LL <- pa.bym$summary.fitted.values[, "0.025quant"]
pa.tr$UL <- pa.bym$summary.fitted.values[, "0.975quant"]

summary(pa.tr[, c("AdjCounts", "LL", "UL", "exc1")])

pa.tr_sf$exc1 = pa.tr$exc1
pa.tr_sf$exc2 = round(pa.tr_sf$exc1,2)
table(pa.tr_sf$exc2)

pa.tr_sf$Exceedance = ifelse(pa.tr_sf$exc1<=0.5,0,
                          ifelse(pa.tr_sf$exc1>0.5,1,NA))


pa.tr_sf$Exceedance <- ordered(pa.tr_sf$Exceedance,
                             levels = c(0,1),
                             labels = c("Less than 50%", "More than 50%"))

table(pa.tr_sf$Exceedance)

plot_usmap(include = "PA", data = pa.tr_sf, values = "Exceedance") + 
  scale_fill_discrete(name = "Exceedance Probability", h= c(0,360) + 5, c=100, l=65, h.start =60, direction=1, na.value="grey50", aesthetics ="fill") +  labs(title = "Exceedance Probability: Counties with Rate Ratios Above 1%") + 
  theme(panel.background = element_rect(colour = "black", fill = "light grey")) + theme(legend.position = "right")


```

#Spatiotemporal Model

```{r}
#let's read the data

covid.data = read.csv("dataCovid_PA.csv", header = T)

head(covid.data)

```


```{r}
april.data = subset(covid.data, select = c(dApr30, Black_Percent, Population, years65, Poverty, somecollegeassociate, Obesity, Metropolitan, fips))
april.data$timeID = 1
april.data$rate <- sum(april.data$dApr30)/sum(april.data$Population)
april.data$EC <- april.data$rate*april.data$Population
names(april.data)[1] <- "count"
april.data$SMR = april.data$count/april.data$EC


may.data = subset(covid.data, select = c(dMay31, Black_Percent, Population, years65, Poverty, somecollegeassociate, Obesity, Metropolitan, fips))
may.data$timeID = 2
may.data$rate <- sum(may.data$dMay31)/sum(may.data$Population)
may.data$EC <- may.data$rate*may.data$Population
names(may.data)[1] <- "count"
may.data$SMR = may.data$count/may.data$EC

june.data = subset(covid.data, select = c(dJun30, Black_Percent, Population, years65, Poverty, somecollegeassociate, Obesity, Metropolitan, fips))
june.data$timeID = 3
june.data$rate <- sum(june.data$dJun30)/sum(june.data$Population)
june.data$EC <- june.data$rate*june.data$Population
names(june.data)[1] <- "count"
june.data$SMR = june.data$count/june.data$EC


july.data = subset(covid.data, select = c(dJul31, Black_Percent, Population, years65, Poverty, somecollegeassociate, Obesity, Metropolitan, fips))
july.data$timeID = 4
july.data$rate <- sum(july.data$dJul31)/sum(july.data$Population)
july.data$EC <- july.data$rate*july.data$Population
names(july.data)[1] <- "count"
july.data$SMR = july.data$count/july.data$EC


august.data = subset(covid.data, select = c(dAug31, Black_Percent, Population, years65, Poverty, somecollegeassociate, Obesity, Metropolitan, fips))
august.data$timeID = 5
august.data$rate <- sum(august.data$dAug31)/sum(august.data$Population)
august.data$EC <- august.data$rate*august.data$Population
names(august.data)[1] <- "count"
august.data$SMR = august.data$count/august.data$EC


```

```{r}
table(august.data$SMR)
mean(august.data$SMR)
```



```{r}
covid.tr = st_as_sf(pa.tr)

aprilcovid= merge(covid.tr, april.data, by = 'fips')

maycovid= merge(covid.tr, may.data, by = 'fips')

junecovid= merge(covid.tr, june.data, by = 'fips')

julycovid= merge(covid.tr, july.data, by = 'fips')

augustcovid= merge(covid.tr, august.data, by = 'fips')
```


```{r}
#April 
ggplot(aprilcovid) + geom_sf(aes(fill = SMR)) +
  scale_fill_gradient2(
    midpoint = 0.5, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()

#May 
ggplot(maycovid) + geom_sf(aes(fill = SMR)) +
  scale_fill_gradient2(
    midpoint = 0.5, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()

#June 
ggplot(junecovid) + geom_sf(aes(fill = SMR)) +
  scale_fill_gradient2(
    midpoint = 0.5, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()

#July 
ggplot(julycovid) + geom_sf(aes(fill = SMR)) +
  scale_fill_gradient2(
    midpoint = 0.5, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()

#August 
ggplot(augustcovid) + geom_sf(aes(fill = SMR)) +
  scale_fill_gradient2(
    midpoint = 0.5, low = "blue", mid = "white", high = "red",
    limits= c(0,2)  ) +
  theme_bw()
```





```{r}

count <- as.vector(c(april.data$count, may.data$count, june.data$count, july.data$count, august.data$count))

black <- as.vector(c(april.data$Black_Percent, may.data$Black_Percent, june.data$Black_Percent, july.data$Black_Percent, august.data$Black_Percent))

population <- as.vector(c(april.data$Population, may.data$Population, june.data$Population, july.data$Population, august.data$Population))

years65 <- as.vector(c(april.data$years65, may.data$years65, june.data$years65, july.data$years65, august.data$years65))

poverty <- as.vector(c(april.data$Poverty, may.data$Poverty, june.data$Poverty, july.data$Poverty, august.data$Poverty))

college <- as.vector(c(april.data$somecollegeassociate, may.data$somecollegeassociate, june.data$somecollegeassociate, july.data$somecollegeassociate, august.data$somecollegeassociate))

obesity <- as.vector(c(april.data$Obesity, may.data$Obesity, june.data$Obesity, july.data$Obesity, august.data$Obesity))

metro <- as.vector(c(april.data$Metropolitan, may.data$Metropolitan, june.data$Metropolitan, july.data$Metropolitan, august.data$Metropolitan))

fips <- as.vector(c(april.data$fips, may.data$fips, june.data$fips, july.data$fips, august.data$fips))

timeID <- as.vector(c(april.data$timeID, may.data$timeID, june.data$timeID, july.data$timeID, august.data$timeID))


smr <- as.vector(c(april.data$SMR, may.data$SMR, june.data$SMR, july.data$SMR, august.data$SMR))

ec <- as.vector(c(april.data$EC, may.data$EC, june.data$EC, july.data$EC, august.data$EC))

data.all <- data.frame(count, black, population, years65, poverty, college, obesity, metro, fips, timeID, smr, ec)



```


```{r}
stformula <- count~ years65 + black + poverty + college + obesity + metro + f(fips, timeID, model = "iid")


nu <- 1:length(count)

prior.fixed <- list(mean.intercept = 0, prec.intercept = 0.0001,
                    mean = 0, prec = 0.0001)
prec.prior <- list(prec = list(param = c(0.001, 0.001)))

Wmat <- W.pa.rs


```


```{r}
model <- inla(
  stformula,                     # Model formula
  data = data.all,               # Data used for modeling
  family = "nbinomial",          # Likelihood family (negative binomial), other families:
  E = EC,                        # Expected counts
  control.fixed = prior.fixed,   # Control options for fixed effects
  control.compute = list(
    waic = T,                    # Compute the Widely Applicable Information Criterion (WAIC)
    return.marginals.predictor = TRUE  # Return marginal effects for predictors
  ),
  control.predictor = list(
    link = 1,                    # Link function (1 for log-link)
    compute = TRUE               # Compute predictions
  )
)

View(exp(model$summary.fixed))
```


